---
title: |
  | Code examples in R for Chapter 7:
  | Modern Portfolio Theory
author: "Daniel P. Palomar"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
header-includes:
  - \newcommand{\bm}[1]{\boldsymbol{#1}}
  - \newcommand{\w}{\bm{w}}
  - \newcommand{\bmu}{\bm{\mu}}
  - \newcommand{\bSigma}{\bm{\Sigma}}
  - \newcommand{\textm}[1]{\textsf{#1}}
  - \def\t{{\mkern-2mu\raise-1mu\mathsf{T}}}    
---

```{r, echo=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "R>>",
  fig.align = "center"
)
options(width=120)
```
------------


# Packages
The following packages are used in the examples:
```{r, echo=TRUE, message=FALSE}
# basic finance
library(xts)                    # to manipulate time series of stock data
library(PerformanceAnalytics)   # to compute performance measures
library(portfolioBacktest)      # to conduct backtests
library(pob)                    # book package with financial data

# plotting
library(ggplot2)                # for nice plots
library(magrittr)               # for pipeline operator %>%
library(reshape2)

# optimization
library(CVXR)
```




# MVP
We explore the mean-variance portfolio (MVP),
$$
  \begin{array}{ll}
  \underset{\w}{\textm{maximize}} & \w^\t\bmu - \frac{\lambda}{2}\w^\t\bSigma\w\\
  \textm{subject to} & \bm{1}^\t\w=1, \quad \w\ge\bm{0},
  \end{array}
$$
with different values of the hyper-parameter $\lambda$.

```{r, echo=TRUE, message=TRUE, fig.width=9, fig.height=4}
data(SP500_2015to2020)
stock_prices <- SP500_2015to2020$stocks["2019::", c("AAPL", "AMZN", "AMD", "GM", "GOOGL", "MGM", "MSFT", "QCOM", "TSCO", "UPS")]
T <- nrow(stock_prices)
T_trn <- round(0.70*T)

#
# Define portfolios
#
EWP <- function(data, ...) {
  N <- ncol(data$prices)
  return(rep(1/N, N))
}

design_MVP <- function(mu, Sigma, lambda = 1, ub = Inf) {
  w <- Variable(nrow(Sigma))
  prob <- Problem(Maximize(t(mu) %*% w - (lambda/2)*quad_form(w, Sigma)),
                  constraints = list(w >= 0, sum(w) == 1, w <= ub))
  result <- solve(prob)
  w <- as.vector(result$getValue(w))
  return(w)
}

MVP_lmd1 <- function(dataset, ...) {
  N <- ncol(dataset$prices)
  X <- diff(log(dataset$prices))[-1]
  mu <- colMeans(X)
  Sigma <- cov(X)
  w <- design_MVP(mu, Sigma, lambda = 1)
  return(w)
}

MVP_lmd4 <- function(dataset, ...) {
  N <- ncol(dataset$prices)
  X <- diff(log(dataset$prices))[-1]
  mu <- colMeans(X)
  Sigma <- cov(X)
  w <- design_MVP(mu, Sigma, lambda = 4)
  return(w)
}

MVP_lmd16 <- function(dataset, ...) {
  N <- ncol(dataset$prices)
  X <- diff(log(dataset$prices))[-1]
  mu <- colMeans(X)
  Sigma <- cov(X)
  w <- design_MVP(mu, Sigma, lambda = 16)
  return(w)
}

MVP_lmd64 <- function(dataset, ...) {
  N <- ncol(dataset$prices)
  X <- diff(log(dataset$prices))[-1]
  mu <- colMeans(X)
  Sigma <- cov(X)
  w <- design_MVP(mu, Sigma, lambda = 64)
  return(w)
}


# single backtest
bt <- portfolioBacktest(portfolio_funs = list("1/N"             = EWP,
                                              "MVP (lmd = 1)"   = MVP_lmd1,
                                              "MVP (lmd = 4)"   = MVP_lmd4,
                                              "MVP (lmd = 16)"  = MVP_lmd16,
                                              "MVP (lmd = 64)"  = MVP_lmd64),
                        dataset_list = list(list("prices" = stock_prices)), price_name = "prices",
                        lookback = T_trn, optimize_every = 30, rebalance_every = 1)

# barplot
data.frame(
  "stocks" = names(stock_prices),
  "1/N"    = as.numeric(bt$`1/N`$data1$w_optimized[1, ]),
  "MVP (lmd = 1)"    = as.numeric(bt$`MVP (lmd = 1)`$data1$w_optimized[1, ]),
  "MVP (lmd = 4)"    = as.numeric(bt$`MVP (lmd = 4)`$data1$w_optimized[1, ]),
  "MVP (lmd = 16)"   = as.numeric(bt$`MVP (lmd = 16)`$data1$w_optimized[1, ]),
  "MVP (lmd = 64)"   = as.numeric(bt$`MVP (lmd = 64)`$data1$w_optimized[1, ]),
  check.names = FALSE
) %>% melt(., id.vars = "stocks") %>%
  ggplot(aes(x = stocks, y = value, fill = variable)) +
  geom_bar(stat = "identity", position = "dodge", color = "black", width = 0.8) +
  labs(title = "Capital weight allocation for different portfolios", y = "weight", fill = "portfolios")
```

```{r, echo=TRUE, fig.width=9, fig.height=4}
backtestChartCumReturn(bt) +
  geom_line(size = 0.8) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %Y", date_minor_breaks = "1 day") +
  ggtitle("Cumulative return (out of sample)")

backtestChartDrawdown(bt) +
  geom_line(size = 0.8) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %Y", date_minor_breaks = "1 day") +
  ggtitle("Drawdown (out of sample)")

summary_bt <- backtestSummary(bt)
summaryTable(summary_bt, type = "DT",
             measures = c("Sharpe ratio", "annual return", "annual volatility", "max drawdown"))
```


##### MVP with diversification heuristics {-}
We now consider the MVP with two diversification heuristics:

- upper bound: $\w\leq0.25\times\bm{1}$; and
- diversification constraint: $\|\w\|_2^2 \leq 2/N$.


```{r, echo=TRUE, message=TRUE, fig.width=9, fig.height=4}
data(SP500_2015to2020)
stock_prices <- SP500_2015to2020$stocks["2019::", c("AAPL", "AMZN", "AMD", "GM", "GOOGL", "MGM", "MSFT", "QCOM", "TSCO", "UPS")]
T <- nrow(stock_prices)
T_trn <- round(0.70*T)

#
# Define portfolios
#
design_GMVP <- function(Sigma) {
  w <- Variable(nrow(Sigma))
  prob <- Problem(Minimize(quad_form(w, Sigma)),
                  constraints = list(w >= 0, sum(w) == 1))
  result <- solve(prob)
  w <- as.vector(result$getValue(w))
  return(w)
}

GMVP <- function(dataset, ...) {
  N <- ncol(dataset$prices)
  X <- diff(log(dataset$prices))[-1]
  Sigma <- cov(X)
  w <- design_GMVP(Sigma)
  return(w)
}

MVP <- function(dataset, ...) {
  N <- ncol(dataset$prices)
  X <- diff(log(dataset$prices))[-1]
  mu <- colMeans(X)
  Sigma <- cov(X)
  w <- design_MVP(mu, Sigma)
  return(w)
}

MVP_ub <- function(dataset, ...) {
  N <- ncol(dataset$prices)
  X <- diff(log(dataset$prices))[-1]
  mu <- colMeans(X)
  Sigma <- cov(X)
  w <- design_MVP(mu, Sigma, ub = 0.25)
  return(w)
}

design_MVP_diversification <- function(mu, Sigma, lmd = 1, D) {
  w <- Variable(nrow(Sigma))
  prob <- Problem(Maximize(t(mu) %*% w - (lmd/2)*quad_form(w, Sigma)),
                  constraints = list(w >= 0, sum(w) == 1, sum(w^2) <= D))
  result <- solve(prob)
  w <- as.vector(result$getValue(w))
  return(w)
}

MVP_diversification <- function(dataset, ...) {
  N <- ncol(dataset$prices)
  X <- diff(log(dataset$prices))[-1]
  mu <- colMeans(X)
  Sigma <- cov(X)
  w <- design_MVP_diversification(mu, Sigma, D = 2/N)
  return(w)
}



# single backtest
bt <- portfolioBacktest(portfolio_funs = list("1/N"   = EWP,
                                              "GMVP"  = GMVP,
                                              "MVP"   = MVP,
                                              "MVP with upper bound"        = MVP_ub,
                                              "MVP with diversific. const." = MVP_diversification),
                        dataset_list = list(list("prices" = stock_prices)), price_name = "prices",
                        lookback = T_trn, optimize_every = 30, rebalance_every = 1)


# barplot
data.frame(
  "stocks" = names(stock_prices),
  "1/N"    = as.numeric(bt$`1/N`$data1$w_optimized[1, ]),
  "GMVP"   = as.numeric(bt$`GMVP`$data1$w_optimized[1, ]),
  "MVP"    = as.numeric(bt$`MVP`$data1$w_optimized[1, ]),
  "MVP with upper bound"        = as.numeric(bt$`MVP with upper bound`$data1$w_optimized[1, ]),
  "MVP with diversific. const." = as.numeric(bt$`MVP with diversific. const.`$data1$w_optimized[1, ]),
  check.names = FALSE
) %>% melt(., id.vars = "stocks") %>%
  ggplot(aes(x = stocks, y = value, fill = variable)) +
  geom_bar(stat = "identity", position = "dodge", color = "black", width = 0.8) +
  labs(title = "Capital weight allocation for different portfolios", y = "weight", fill = "portfolios")
```

```{r, echo=TRUE, fig.width=9, fig.height=4}
backtestChartCumReturn(bt) +
  geom_line(size = 0.8) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %Y", date_minor_breaks = "1 day") +
  ggtitle("Cumulative return (out of sample)")

backtestChartDrawdown(bt) +
  geom_line(size = 0.8) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %Y", date_minor_breaks = "1 day") +
  ggtitle("Drawdown (out of sample)")

summary_bt <- backtestSummary(bt)
summaryTable(summary_bt, type = "DT",
             measures = c("Sharpe ratio", "annual return", "annual volatility", "max drawdown"))
```




# MSRP
We now compare the following portfolios:

- global minimum variance portfolio (GMVP):
$$
\begin{array}{ll}
\underset{\w}{\textm{minimize}} & \w^\t\bSigma\w\\
\textm{subject to} & \bm{1}^\t\w=1, \quad \w\ge\bm{0};
\end{array}
$$

- mean-variance portfolio (MVP):
$$
  \begin{array}{ll}
  \underset{\w}{\textm{maximize}} & \w^\t\bmu - \frac{\lambda}{2}\w^\t\bSigma\w\\
  \textm{subject to} & \bm{1}^\t\w=1, \quad \w\ge\bm{0};
  \end{array}
$$

- maximum Sharpe ratio portfolio (MSRP):
$$
  \begin{array}{ll}
  \underset{\w}{\textm{maximize}} & \dfrac{\w^\t\bmu - r_\textm{f}}{\sqrt{\w^\t\bSigma\w}}\\
  \textm{subject to} & \begin{array}{l} \bm{1}^\t\w=1, \quad \w\ge\bm{0}.\end{array}
  \end{array}
$$

```{r, echo=TRUE, message=TRUE, fig.width=9, fig.height=4}
data(SP500_2015to2020)
stock_prices <- SP500_2015to2020$stocks["2019::", c("AAPL", "AMZN", "AMD", "GM", "GOOGL", "MGM", "MSFT", "QCOM", "TSCO", "UPS")]
T <- nrow(stock_prices)
T_trn <- round(0.70*T)

#
# Define portfolios
#
GMVP <- function(dataset, ...) {
  N <- ncol(dataset$prices)
  X <- diff(log(dataset$prices))[-1]
  Sigma <- cov(X)
  w <- design_GMVP(Sigma)
  return(w)
}

MVP <- function(dataset, ...) {
  N <- ncol(dataset$prices)
  X <- diff(log(dataset$prices))[-1]
  mu <- colMeans(X)
  Sigma <- cov(X)
  w <- design_MVP(mu, Sigma, lambda = 4)
  return(w)
}


design_MSRP <- function(mu, Sigma) {
  w_ <- Variable(nrow(Sigma))
  prob <- Problem(Minimize(quad_form(w_, Sigma)),
                  constraints = list(w_ >= 0, t(mu) %*% w_ == 1))
  result <- solve(prob)
  w <- as.vector(result$getValue(w_)/sum(result$getValue(w_)))
  names(w) <- colnames(Sigma)
  return(w)
}

MSRP <- function(dataset, ...) {
  N <- ncol(dataset$prices)
  X <- diff(log(dataset$prices))[-1]
  mu <- colMeans(X)
  Sigma <- cov(X)
  w <- design_MSRP(mu, Sigma)
  return(w)
}


# single backtest
bt <- portfolioBacktest(portfolio_funs = list("GMVP"  = GMVP,
                                              "MVP"   = MVP,
                                              "MSRP"  = MSRP),
                        dataset_list = list(list("prices" = stock_prices)), price_name = "prices",
                        lookback = T_trn, optimize_every = 30, rebalance_every = 1)

# barplot
data.frame(
  "stocks" = names(stock_prices),
  "GMVP"   = as.numeric(bt$`GMVP`$data1$w_optimized[1, ]),
  "MVP"    = as.numeric(bt$`MVP`$data1$w_optimized[1, ]),
  "MSRP"    = as.numeric(bt$`MSRP`$data1$w_optimized[1, ]),
  check.names = FALSE
) %>% melt(., id.vars = "stocks") %>%
  ggplot(aes(x = stocks, y = value, fill = variable)) +
  geom_bar(stat = "identity", position = "dodge", color = "black", width = 0.8) +
  labs(title = "Capital weight allocation for different portfolios", y = "weight", fill = "portfolios")
```

```{r, echo=TRUE, fig.width=9, fig.height=4}
backtestChartCumReturn(bt) +
  geom_line(size = 0.8) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %Y", date_minor_breaks = "1 day") +
  ggtitle("Cumulative return (out of sample)")

backtestChartDrawdown(bt) +
  geom_line(size = 0.8) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %Y", date_minor_breaks = "1 day") +
  ggtitle("Drawdown (out of sample)")

summary_bt <- backtestSummary(bt)
summaryTable(summary_bt, type = "DT",
             measures = c("Sharpe ratio", "annual return", "annual volatility", "max drawdown"))
```




# Universal algorithm
We consider two methods for the resolution of the MSRP:

- via the Schaible transform, i.e., solving
$$
  \begin{array}{ll}
  \underset{\bm{y}}{\textm{minimize}} & \bm{y}^\t\bSigma\bm{y}\\
  \textm{subject to} 
  & \bm{y}^\t\left(\bmu - r_\textm{f}\bm{1}\right) = 1\\
  & \bm{y}\ge\bm{0},
  \end{array}
$$
with $\w = \bm{y}/\left(\bm{1}^\t\bm{y}\right)$, and then 

- via the universal iterative SQP-MVP algorithm, i.e., we iteratively obtain $\w^{k+1}$, for $k=0,1,\dots,$, by solving 
$$
  \begin{array}{ll}
  \underset{\w}{\textm{maximize}} & \w^\t\bmu - \dfrac{\lambda^k}{2}\w^\t\bSigma\w\\
  \textm{subject to} & \bm{1}^\t\w=1, \quad \w\ge\bm{0},
  \end{array}
$$
with
$$
  \lambda^k = \dfrac{(\w^k)^\t\bmu - r_\textm{f}}{(\w^k)^\t\bSigma\w^k}.
$$


```{r, echo=TRUE, message=TRUE, fig.width=9, fig.height=4}
data(SP500_2015to2020)
stock_prices <- SP500_2015to2020$stocks["2019::", c("AAPL", "AMZN", "AMD", "GM", "GOOGL", "MGM", "MSFT", "QCOM", "TSCO", "UPS")]
X <- diff(log(stock_prices))[-1]
T <- nrow(X)
N <- ncol(X)
mu <- colMeans(X)
Sigma <- cov(X)

# using Schaible transform
w_ <- Variable(N)
prob <- Problem(Minimize(quad_form(w_, Sigma)),
                constraints = list(w_ >= 0, t(mu) %*% w_ == 1))
result <- solve(prob)
w_cvx <- as.vector(result$getValue(w_)/sum(result$getValue(w_)))
obj_cvx <- as.numeric(t(w_cvx) %*% mu / sqrt(t(w_cvx) %*% Sigma %*% w_cvx))

# using the SQP-MVP algorithm
w <- rep(1/N, N)
obj_sqp <- c(t(w) %*% mu / sqrt(t(w) %*% Sigma %*% w))
k <- 0
for (k in 0:20) {
  lmd_k <- as.numeric(t(w) %*% mu / (t(w) %*% Sigma %*% w))
  w_prev <- w
  w <- Variable(N)
  prob <- Problem(Maximize(t(w) %*% mu - (lmd_k/2) * quad_form(w, Sigma)),
                  constraints = list(sum(w) == 1, w >= 0))
  result <- solve(prob)
  w <- as.vector(result$getValue(w))
  obj_sqp <- c(obj_sqp, t(w) %*% mu / sqrt(t(w) %*% Sigma %*% w))
  k <- k + 1
  if (max(abs(w - w_prev)) < 1e-4)
    break
}

# plot
data.frame(
  "k"                     = c(0:k),
  "SQP iterative method"  = obj_sqp,
  check.names = FALSE
) %>% melt(., id.vars = "k") %>%
  ggplot(aes(x = k, y = value, color = variable, fill = variable)) +
  geom_hline(yintercept = obj_cvx, size = 0.8) +
  geom_line(size = 0.8, show.legend = FALSE) + 
  geom_point(size = 1.5, show.legend = FALSE) +
  ggtitle("Convergence") + ylab("objective value") +  xlab("iteration")
```



Then consider the resolution of the mean-volatility portfolio via an SOCP solver and then via the universal SQP-MVP algorithm:
```{r, echo=TRUE, message=TRUE, fig.width=9, fig.height=4}
data(SP500_2015to2020)
stock_prices <- SP500_2015to2020$stocks["2019::", c("AAPL", "AMZN", "AMD", "GM", "GOOGL", "MGM", "MSFT", "QCOM", "TSCO", "UPS")]
X <- diff(log(stock_prices))[-1]
T <- nrow(X)
N <- ncol(X)
mu <- colMeans(X)
Sigma <- cov(X)
kappa <- 1.0

# using CVX
Sigma_12 <- chol(Sigma)
#max(abs(t(Sigma_12) %*% Sigma_12 - Sigma))  # sanity check
w <- Variable(N)
prob <- Problem(Maximize(t(w) %*% mu - kappa*cvxr_norm(Sigma_12 %*% w, 2)),
                constraints = list(sum(w) == 1, w >= 0))
result <- solve(prob)  #result$status
w_cvx <- as.vector(result$getValue(w))
obj_cvx <- as.numeric(t(w_cvx) %*% mu - kappa*sqrt(t(w_cvx) %*% Sigma %*% w_cvx))  # same as: result$value

# using the SQP-MVP algorithm
w <- rep(1/N, N)
obj_sqp <- c(t(w) %*% mu - sqrt(t(w) %*% Sigma %*% w))
k <- 0
for (k in 0:20) {
  lmd_k <- kappa/sqrt(t(w) %*% Sigma %*% w)
  w_prev <- w
  w <- Variable(N)
  prob <- Problem(Maximize(t(w) %*% mu - (lmd_k/2) * quad_form(w, Sigma)),
                  constraints = list(sum(w) == 1, w >= 0))
  result <- solve(prob)
  w <- as.vector(result$getValue(w))
  obj_sqp <- c(obj_sqp, t(w) %*% mu - kappa*sqrt(t(w) %*% Sigma %*% w))
  k <- k + 1
  if (max(abs(w - w_prev)) < 1e-4)
    break
}

# plot
data.frame(
  "k"                     = c(0:k),
  #"SOCP solver"           = rep(obj_cvx, k+1),
  "SQP iterative method"  = obj_sqp,
  check.names = FALSE
) %>% melt(., id.vars = "k") %>%
  ggplot(aes(x = k, y = value, color = variable, fill = variable)) +
  geom_hline(yintercept = obj_cvx, size = 0.8) +
  geom_line(size = 0.8, show.legend = FALSE) + 
  geom_point(size = 1.5, show.legend = FALSE) +
  ggtitle("Convergence") + ylab("objective value") +  xlab("iteration")
```


