---
output:
  html_document:
    variant: markdown_github
    keep_md: true
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.align = "center",
  fig.retina = 2,
  out.width = "75%",
  dpi = 96
)
knit_hooks$set(pngquant = hook_pngquant)
```

# pob
This is a supporting package for the [Portfolio Optimization](https://www.danielppalomar.com/teaching-courses/fall-2021-22-mafs5310-portfolio-optimization-with-r/) course notes.

To install in R:
```{r, eval=FALSE}
devtools::install_github("dppalomar/pob")
```

To install in Python:
```{r, eval=FALSE}
# TBD
```


# Data
The package contains stock data and cryptocurrency data:
```{r}
library(pob)

# stock S&P500 market data
data(SP500_2015to2020)
names(SP500_2015to2020)
head(SP500_2015to2020$stocks[, 1:5])

# crypto data
data(cryptos_2017to2021)
names(cryptos_2017to2021)
head(cryptos_2017to2021$hourly[, 1:5])
```



# Code examples

- Chapter 6: Portfolio Basics
  + code examples in R: [Rmd](code_examples/R/chapter-6/Rsession.Rmd), [html](https://raw.githack.com/dppalomar/pob/master/code_examples/R/chapter-6/Rsession.html), 
  + code examples in Python: TBD
  
- Chapter 7: Modern Portfolio Theory
  + code examples in R: TBD
  + code examples in Python: TBD
  
  
  
